Measuring Stability and Change in Personal Culture Using Panel Data

### 2019-12-06 old_drafts/manuscript8.Rmd
kevin's revisions

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to major theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another (CITE???). Separate from this distinction is a second dimension of attitude formation that reflects degree to which people can consistently report the same opinion over time. Opinions can be settled but difficult to report, such as when individuals construct responses to the same question over time from a settled but diverse set of social cues [@zaller]. Attitudes can also be in a process of active updating but relatively easy to report, such as when a voter changes her registered partisan affiliation.

@@

With these two dimensions outlined, we can begin to categorize predictions of existing theories of attitude development with respect to them. Table ## sketches these two dimensions and outlines where a number of theoretical propositions align in this matrix. We consider these theories in more detail below but note at the outset that our goal here is not to make an explicit prediction for the kind of attitude change we will see for all kinds of attitudes. Instead, we seek to identify sources of tension in these existing models and provide a lens through which to interpret the overall patterns of attitude change across a wide range of types of attitudes.

\begin{center}
\begin{tabular}{ | c | p{4cm} | p{4cm} | } 
\hline
 & Low Active Updating & High Active Updating \\ 
\hline
High Consistency & Attitudes shaped by cohort effects (older individuals); High-profile, settled public policy questions; Core beliefs & High-profile, changing public policy questions; Public attitudes and behaviors; "Rational" beliefs and behaviors \\ 
\hline
Low Consistency &  \\ 
\hline
\end{tabular}
\end{center}

@@

In broad terms, cohort replacement theories of public opinion change posit that individuals hold relatively stable opinions, that few individuals change their attitudes as they age, and that most change in public opinion must come from older individuals dying and being replaced in the population with young individuals who hold different opinions [@mannheim; @ryder; @vaiseylizardo]. Implied in these models is the notion that individuals have a period of "cohortization" in which they form these initial beliefs.

@@

There is abundant evidence that adolescence and early adulthood are periods of heightened attitude change, with this change diminishing rapidly or gradually, depending on the attitude [@glenn; @inglehartbaker; @visserkrosnick]. Most theories agree that middle age (40-60) is a time of reduced attitude change, but some also posit that older individuals might also have increased susceptibility to attitude change [@visserkrosnick; @danigelis].

Many of these studies focus on common measures of political attitudes such as ideological and partisan identification [@krosnickalwin], but Vaisey and Lizardo [-@vaiseylizardo] suggest that this pattern should be the case for a broad range of general attitudes about race and gender. Danigelis, Hardy, and Cutler [-@danigelis] also provide evidence that attitudes about economic gender equality and structural reasons for racial inequality see higher levels of "intracohort" change, a good indicator that some cohort members are changing their minds, in younger cohorts than older cohorts. Theories and evidence of secularization also suggest that most change in religious attitudes and behaviors comes through cohort replacement, rather than individuals changing their beliefs throughout their lives [@voaschaves].

@@

@@

That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response. If change is persisting (i.e., if our metaphorical traveler wakes up in the same place she went to sleep in last night), then our best guess is that she will be close to where we saw her last, and previous responses will provide no additional predictive power.

@@

@@

>>>>

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to major theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another. Separate from this distinction is a second dimension of attitude formation that reflects degree to which people can consistently report the same opinion over time. Opinions can be settled but difficult to report, such as when individuals construct responses to the same question over time from a settled but diverse set of social cues [@zaller]. Attitudes can also be in a process of active updating but relatively easy to report, such as when a voter changes her registered partisan affiliation.

@@

With these two dimensions outlined, we can begin to categorize predictions of existing theories of attitude development with respect to them. Table 1 sketches these two dimensions and outlines where a number of theoretical propositions align in this matrix. We consider these theories in more detail below but note at the outset that our goal here is not to make an explicit prediction for the kind of attitude change we will see for all kinds of attitudes. Instead, we seek to identify sources of tension in these existing models and provide a lens through which to interpret the overall patterns of attitude change across a wide range of types of attitudes.

![Table 1: Predicted change patterns for theoretical models](figures/theory_table.png)

@@

In broad terms, cohort replacement theories of public opinion change posit that people hold relatively stable opinions, that few people change their attitudes as they age, and that most change in public opinion must come from older people dying and being replaced in the population with young people who hold different opinions [@mannheim; @ryder; @vaiseylizardo]. Implied in these models is the notion that people have a period of "cohortization" in which they form these initial beliefs.

@@

There is abundant evidence that adolescence and early adulthood are periods of heightened attitude change, with this change diminishing rapidly or gradually, depending on the attitude [@glenn; @inglehartbaker; @visserkrosnick]. Most theories agree that middle age (40-60) is a time of reduced attitude change, but some also posit that older people might also have increased susceptibility to attitude change [@visserkrosnick; @danigelis].

Many of these studies focus on common measures of political attitudes such as ideological and partisan identification [@krosnickalwin], but Vaisey and Lizardo [-@vaiseylizardo] suggest that this pattern should be the case for a broad range of general attitudes about race and gender. Danigelis, Hardy, and Cutler [-@danigelis] also provide evidence that attitudes about economic gender equality and structural reasons for racial inequality see higher levels of "intracohort" change, a good indicator that some cohort members are changing their minds, in younger cohorts than older cohorts. Theories and evidence of secularization also suggest that most change in religious attitudes and behaviors comes through cohort replacement, rather than people changing their beliefs throughout their lives [@voaschaves].

@@

### Panel Conditioning Bias

Another possibility is that it is not everyday experience that drives changes and stability in beliefs, rather that participation in the survey itself produces change (or stability). This is commonly referred to as panel conditioning bias [@warrenhalpernmanners; @ohetal]. While Warren and Halpern-Manners [-@warrenhalpernmanners] outline several forms of panel conditioning, we can group these into two broad patterns. One set of patterns suggests that people’s responses become more consistent over time as participation in the survey forces them to crystalize their beliefs, seek out new information that helps them form beliefs, realize their beliefs are out of sync with the general population, or learn to "game" the survey to get through it faster. If this were taking place in the GSS, it would result in a pattern of high active updating and high consistency, as individuals change between waves 1 and 2, and wave 2 becomes a better predictor of wave 3.

We do not view this kind of panel conditioning as a problem for our theoretical models. If people change their attitudes or behavior or how they respond to survey questions as a result of participating in a survey, they are conforming to the theoretical active updating model, being open to change throughout their life course, and the source of that change is irrelevant. This might lead to a higher estimate of active updating than we might observe in the population, which would hinder our ability to extrapolate our findings, but would still provide evidence that people update their beliefs over time.

A second form of panel conditioning bias would posit that people exhibit low levels of updating because of commitment bias, or an attempt to maintain consistency in their responses over time. In this scenario, individuals respond to a question at wave 1 and give the same response in subsequent waves, even if their true beliefs or positions change. This would be problematic for our study, as it would under-estimate the amount of real change in the population. However, if this is the case, we should observe no active updating in responses and high levels of consistency in response, as it would be illogical for individuals to report random changes if they were attempting to maintain consistency.

@@

That is, the AUM predicts that a respondent's most recent response is the best available predictor of her next response, and that wave 1 carries no additional information in predicting wave 3 once we control for wave 2 (the coefficient on this term should be 0). If change is persisting (i.e., if our metaphorical traveler wakes up in the same place she went to sleep in last night), then our best guess is that she will be close to where we saw her last, and previous responses will provide no additional predictive power.

@@

### Comparison to Other Approaches

Our model is not the first approach to measuring stability and change in panel data, but existing models make assumptions that eliminate the distinction between data-generating processes we seek to test. For example, Hout and Hastings [-@houthastings] use a hierarchical model to measure reliability in GSS survey responses over time. This model assumes that there is no change in the underlying latent item other than wave-specific period effects (akin to our settled dispositions model), so the design precludes the possibility of quantifying the level of active updating in an item over time, assuming that this change is just measurement error. These authors also test a structural-equation model designed by Alwin [-@alwin] and Heise [-@heise] that assumes the process that generates the data is the Markovian active updating process we outline earlier. While this approach gets closer to our model by generating a parameter for stability and reliability, it would require us to make an assumption that the amount of "structural" or "non-structural" change is consistent across waves. This approach combines persisting and non-persisting change into two similar but distinct kinds of change "structural" and "non-structural," both of which can be persisting and non-persisting. This distinction, while important for some theoretical questions, is not our focus.

A number of other approaches seek to understand the consistency of latent beliefs by combining and scaling responses to questions that represent the same latent concept [@ansolabehereetal], assuming that wave-to-wave changes in responses tend to represent measurement errors around a "true" latent belief. This raises the distinction between the _stability of a belief_ and the _stability of a survey question response_. Since we at times invoke both these models, we include a handful of composite scales (discussed below) of related items. If wave-to-wave changes in survey responses are non-persisting measurement errors, then scales should have higher consistency than the measures they comprise, but we should see no difference in persistence.

Finally, a nubmer of approaches exist for evaluating theoretical process of belief change for the population, such as examining the association between theoretically related values [@baldassarigelman; @boutylinevaisey], or by looking at changes in the distribution of responses over time [@dimaggio_polarization]. These tools are well designed to address the questions they set out to. However, since our theoretical questions focus on the process of belief change within individuals, these do not speak to our core concerns.

@@

We said at the outset that patterns of change might be the result of panel conditioning, or that the process of participating in the survey leads to more active updating or stability than might be expected in the absence of survey participation. One could view our results through this prism and claim that items that exhibit high active updating and high consistency (such as views on gay marriage or partisan identification) do so because of panel conditioning bias, or that items that exhibit low active updating and high consistency (such as views on abortion or the legalization of marijuana) do so because of commitment bias, but it becomes difficult to explain why these biases operate for specific questions and not others.

We believe the overall pattern of results we observe is more consistent with other theoretical models of belief change than those outlined by panel conditioning. We see too much inconsistency in responses to most questions for models of commitment bias to be applicable. Items that we have theoretical reasons to suspect might succumb to the updating form of panel conditioning bias, such as questions where “respondents’ initial attitudes are less crystallized” [@warrenhalpernmanners: p. 499], questions that “increase respondents’ knowledge of the behavior and/or their motivation to engage in it” [@warrenhalpernmanners: p. 500], or questions that “induce respondents to provide socially nonnormative or stigmatized responses” [@warrenhalpernmanners: p. 501], tend to show low active updating and low consistency.

We cannot and would not want to rule out the possibility that panel conditioning is taking place in the GSS, and we believe it would be worthwhile to explore these same GSS panels for evidence of panel conditioning. However, we do not believe that panel conditioning bias is the principal driver of the overall pattern of change and consistency we observe.
